---
title: "Synthetic Sales Data Analysis Documentation"
author: "Tsegeweyni"
date: "2026-08-04"
output: html_document
---

> **AI Assistance Declaration:** I used ChatGPT to help organize the R Markdown structure, suggest headings, and improve code chunk formatting. Prompts used are documented in the Appendix. I verified the output by running each code section in Posit Cloud and checking the dataset, summaries, and chart. All final decisions and edits were completed by me. I am responsible for the accuracy and originality of this work.

# Purpose

This R Markdown document explains the `sales_analysis.R` script. The script analyzes a synthetic sales dataset, calculates revenue, summarizes sales performance by product and region, identifies the highest-performing product, and creates a bar chart.

# Inputs

The analysis uses the following input file:

- `data/synthetic_sales.csv`

The dataset contains these variables:

- `date`
- `product`
- `region`
- `units_sold`
- `unit_price`

# Required Packages

The project uses the following R packages:

- `readr`
- `dplyr`
- `ggplot2`

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
```

# Import the Dataset

The following code imports the synthetic sales data.

```{r import-data}
sales_data <- read_csv("data/synthetic_sales.csv")
```

# Preview the Dataset

The first six rows are displayed to confirm that the data imported correctly.

```{r preview-data}
head(sales_data)
```

# Inspect the Data Structure

The `str()` function shows the number of observations, variables, and data types.

```{r inspect-data}
str(sales_data)
```

# Calculate Revenue

Revenue is calculated by multiplying `units_sold` by `unit_price`.

```{r calculate-revenue}
sales_data <- sales_data %>%
  mutate(revenue = units_sold * unit_price)

head(sales_data)
```

# Revenue by Product

The following code groups the data by product and calculates total revenue.

```{r product-summary}
product_summary <- sales_data %>%
  group_by(product) %>%
  summarise(
    total_revenue = sum(revenue),
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue))

product_summary
```

# Revenue by Region

The following code groups the data by region and calculates total revenue.

```{r region-summary}
region_summary <- sales_data %>%
  group_by(region) %>%
  summarise(
    total_revenue = sum(revenue),
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue))

region_summary
```

# Highest-Performing Product

The highest-performing product is selected based on total revenue.

```{r highest-product}
highest_product <- product_summary %>%
  slice_max(
    order_by = total_revenue,
    n = 1,
    with_ties = FALSE
  )

highest_product
```

# Sales Visualization

The following bar chart compares total revenue by product.

```{r sales-chart}
ggplot(
  product_summary,
  aes(x = reorder(product, total_revenue), y = total_revenue)
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Total Revenue by Product",
    x = "Product",
    y = "Total Revenue"
  ) +
  theme_minimal()
```

# Outputs

The analysis produces:

- A sales dataset with a calculated `revenue` variable
- Total revenue by product
- Total revenue by region
- The highest-performing product
- A bar chart of product revenue

# Verification

I ran each code chunk in Posit Cloud and confirmed that:

- The CSV file imported successfully.
- The dataset contained 12 observations and 5 original variables.
- The revenue column was calculated correctly.
- Product and region summaries were created.
- Laptop was identified as the highest-performing product.
- The bar chart rendered correctly.